Simultaneous segmentation and registration of diffusion MR images of the brain driven by active-contours

### 2015-02-20 Publications/2015-NeuroImage/04-methods.tex
discussion refactoring

  that is the dual expression to the energy functional corresponding
  to a discrete \gls*{acwe} framework \citep{chan_active_2001}
  with anisotropic regularization as studied in
  \citep{nagel_investigation_1986}.
We provide a proof of duality in {\color{red} [REF figshare?]}.

@@

{\color{red} \paragraph*{Evidencing convergence and settings}
discuss choice of $\tau$, Courant-Friedrichs-Lewy (CFL) condition / Wolfe conditions,
multi-resolutions on the bspline grid, image samples and surface sampling,  etc.
}

@@

\paragraph*{Real datasets} %
\label{sec:human_connectome}
%
For the evaluation of the algorithm on real \gls*{dmri} data of human brains,
  we collected 15 datasets from the ``minimally preprocessed''
	 database of the \gls*{hcp}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend a large of images, containing \gls*{t1}, \gls*{t2} and
  multi-shell \gls*{dmri} images.
The original acquisitions are released within ``unprocessed'' packages, and
  the ``minimally preprocessed'' are corrected for artifacts, brain-extracted
  and spatially normalized, along with some results of the standard processing with
  \emph{FreeSurfer}.

Selecting the appropriate labels in the \emph{aparc} segmentation, we applied
  the marching-cubes algorithm again to extract the surface of the following
  six homogeneous regions $\Omega_l$:
  1) the thalamus;
  2) \gls*{csf} of the ventricular system and deep \gls*{gm} structures;
  3) cerebral \gls*{wm};
  4) brain stem and cerebellar \gls*{wm};
	5) cerebellar \gls*{gm}; and
	6) cortical \gls*{gm} surface.
The choice of this particular model of homogeneous regions responds to the joint distribution
  of \gls*{fa} and \gls*{adc} of the human brain, as shown in the {\color{red} Supplemental
  Materials (Figure XX)}.

>>>>

  that is the dual expression to the energy functional corresponding
  to a discrete \gls*{acwe} framework \citep{chan_active_2001}
  with anisotropic regularization as studied in
  \citep{nagel_investigation_1986}.

@@

\paragraph*{Settings, implementation details, and evidencing convergence}
\label{sec:conv_report}
The registration parameters (such as $\tau$, the multi-resolutions on the B-Spline grid,
  target image smoothing, etc.) and implementation details (for instance the sparse matrix approach
  to fast interpolation) are dicussed in the {\color{red} Supplemental Materials, section XX.XX}.
The actual choice of parameter settings are publicly distributed with the source code of the experiments.
Additionally, we release our software along with a tool for generating convergence reports to
  demonstrate the behavior of the tool and help scientists configure their experiments.
A sample of report is found in the {\color{red} Supplemental Materials, section XX.XX}.

@@

\paragraph*{Real datasets} %
\label{sec:human_connectome}
%
For the evaluation of the algorithm on real \gls*{dmri} data of human brains,
  we collected 16 datasets from the ``minimally preprocessed''
	 database of the \gls*{hcp}.
We refer the reader to \citep{essen_human_2012} for exact details about acquisition
  parameters, and \citep{glasser_minimal_2013} for the preprocessing issues.
The datasets comprehend a large of images, containing \gls*{t1}, \gls*{t2} and
  multi-shell \gls*{dmri} images.
The original acquisitions are released within ``unprocessed'' packages, and
  the ``minimally preprocessed'' are corrected for artifacts, brain-extracted
  and spatially normalized, along with some results of the standard processing with
  \emph{FreeSurfer}.

Selecting the appropriate labels in the \emph{aparc} segmentation, we applied
  the marching-cubes algorithm again to extract the surface of the following
  six homogeneous regions $\Omega_l$:
  1) the thalamus;
  2) \gls*{csf} of the ventricular system and deep \gls*{gm} structures;
  3) cerebral \gls*{wm};
  4) brain stem and cerebellar \gls*{wm};
	5) cerebellar \gls*{gm}; and
	6) cortical \gls*{gm} surface.
The choice of this particular model is further addressed in \nameref{sec:res_model_and_metric}.

### 2015-02-20 Publications/2015-NeuroImage/05-results.tex
discussion refactoring

\subsection*{Proof of concept on digital phantoms}
\label{sec:results_phantom}
Using the experimental instrumentation described in section \nameref{sec:experiments_evaluation},
  we collected quantitative results on the digital phantoms.
To warp the simulated datasets, we randomly generated 150 different realizations of
  the distortion $U^{-1}_{true}$.
Smoothness is introduced using two levels of B-Spline functions, with control points evenly
  located in isotropic grids covering the full extent of the phantom.
The first level had $50.5mm$ of separation between control points and the second $25.25mm$.
The rationale behind this choice is producing large deformations on the contours with
  the first level, and matching the properties of susceptibility-derived distortions
  as previously reported by \cite{irfanoglu_susceptibility_2011} with the finer grid.
Invertibility of $U^{-1}_{true}$ is ensured by controlling the maximum displacement at each
  level ($20.2mm$ and $10.1mm$ respectively) as studied in \citep{rueckert_diffeomorphic_2006}.

A total of 1200 registration experiments (4 phantom types $\times$ 150 random warpings
  $\times$ 2 resolutions) were run.
For each registration, the averaged Hausdorff distance (see \nameref{sec:experiments_evaluation})
  of the inner and the outer surfaces was computed.
The results showed a consistent and high accuracy, below voxel resolution.
\autoref{fig:phantom} (block C) presents the box plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the missregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel resolution with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).

@@

\paragraph*{Assessing the segmentation model}\label{sec:res_model_and_metric} %
%
Preliminarly, we investigated the aptness of our segmentation model to play as cost function
  driving the registration process.
First, we selected the 6-regions model described in \nameref{sec:human_connectome} after
  visual evaluation of the joint distribution {\color{red} REF Supplemental} of several
  alternative models for segmenting the target bivariate images (\gls*{fa} and \gls*{adc}).
Since susceptibility-derived distortions are always aligned with the phase encoding axis
  and feature the property of being linear with respect to the phase inhomogeneity map,
  we reduced the dimensionality of the error space to one dimension.
Thus, the energy functional \eqref{eq:energy} was evaluated for the contour positions
  obtained along a lineralized space of distortions
  $\hat{U}\{\vec{r}\} = \epsilon \cdot U_{true} = \vec{r} + \epsilon \cdot u(\vec{r})$,
  with $\epsilon \in [-1.1, 1.1]$.
The result of this experiment is shown in \autoref{fig:energymap}.
The metric consistently displayed its minimum at the zero-error point.

@@

\paragraph*{Cross-comparison evaluation}\label{sec:res_cc_evaluation}
%
Finally, we evaluated the performance of our registration method, and compared it to correcting
  for distortions using the \emph{b0}-to-\gls*{t2} registration approach.

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/figure05.pdf}
	\caption{Results on real data}\label{fig:results_real}
\end{figure*}

>>>>

\subsection*{Proof of concept on digital phantoms}
\label{sec:results_phantom}
Using a simplified subset of the experimental instrumentation described in section
  \nameref{sec:experiments_evaluation}, we collected quantitative results on the digital
  phantoms.
To warp the simulated datasets, we randomly generated 150 different realizations of
  the distortion $U^{-1}_{true}$.
Smoothness is introduced using two levels of B-Spline functions, with control points evenly
  located in isotropic grids covering the full extent of the phantom.
The first level had $50.5mm$ of separation between control points and the second $25.25mm$.
The rationale behind this choice is producing large deformations on the contours with
  the first level, and matching the properties of susceptibility-derived distortions
  as previously reported by \cite{irfanoglu_susceptibility_2011} with the finer grid.
Invertibility of $U^{-1}_{true}$ is ensured by controlling the maximum displacement at each
  level ($20.2mm$ and $10.1mm$ respectively) as studied in \citep{rueckert_diffeomorphic_2006}.

A total of 1200 registration experiments (4 phantom types $\times$ 150 random warpings
  $\times$ 2 resolutions) were run.
Parameter settings of every processing step in the experiments are made available along with
  the source code.
For each registration, the averaged Hausdorff distance (see \nameref{sec:experiments_evaluation})
  of the inner and the outer surfaces was computed.
The results showed a consistent and high accuracy, below voxel resolution.
\autoref{fig:phantom} (block C) presents the box plots by model type corresponding
  to the two sets of resolutions of generated phantoms.
To support that the missregistration errors averaged per experiment was significantly
  under the resolution of the target image, we proceeded as follows.
First, we confirmed that the error distributions were skewed using a Shapiro-Wilk test of
  normality.
All the distributions of errors under test (4 phantom types $\times$ 2 resolutions) resulted
  non-normal with $p<0.001$.
Consequently, we used the non-parametric Wilcoxon signed-rank test along with Bonferroni
  correction for multiple comparisons ($N=150$).
Averaged errors resulted significantly lower than voxel resolution with $p < (0.001 / 150)$
  for all the tests (4 phantom types $\times$ 2 resolutions).

@@

\paragraph*{Assessing the segmentation model}\label{sec:res_model_and_metric} %
%
Preliminarly, we investigated the aptness of our segmentation model to play as cost function
  driving the registration process.
First, we evaluated several alternative models to segment the target bivariate images
  (\gls*{fa} and \gls*{adc}).
Based on the observed joint distribution of each (see {\color{red} Supplemental Materials, ...}),
  we selected the 6-regions scheme described in \nameref{sec:human_connectome}.

The distortions follow expression \eqref{eq:fieldmap} and hence, we easily reduced the
  error space to one dimension as follows.
The energy functional \eqref{eq:energy} was evaluated for the warpings given by
  a lineralized space of distortions
  $\hat{U}\{\vec{r}\} = \epsilon \cdot U_{true} = \vec{r} + \epsilon \cdot u(\vec{r})$,
  with $\epsilon \in [-1.1, 1.1]$.
The result of this experiment is shown in \autoref{fig:energymap}.
The metric consistently displayed its minimum at $\epsilon=0.0$ (ground-truth location).

@@

\paragraph*{Evaluation \& cross-comparison}\label{sec:res_cc_evaluation}
%
Finally, we tested the performance of our registration method, and compared it to a competing
  method: the \emph{b0}-to-\gls*{t2} registration approach.
To this end, we integrated the registration software \emph{elastix} \citep{klein_elastix_2010},
  and reproduced the settings that internally uses \emph{ExploreDTI}
  \citep{leemans_exploredti_2009}, a widely used toolkit for tractography analysis of
  \gls*{dti}.
A dual workflow to the general evaluation framework (\autoref{fig:evworkflows})
  was also implemented to run the correction experiment with the competing method.
First, the average \emph{b0} volume is computed using all the \emph{low-b} volumes within
  the distorted \gls*{dmri} dataset.
Using the ``preprocessed'' instance of the \gls*{t2} image of the corresponding subject as
  reference image, and the \emph{b0} image as moving image, we performed the same correction
  experiment.

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/figure05.pdf}
	\caption{The evaluation instrument automatically renders the overlay of the
	  target \gls*{fa} image and the resulting contours after distortion correction (yellow color) and
	  their theoretical location using the ground-truth (green color).
	The first two rows show several axial slices for \emph{regseg} (``proposed'' row) and the
	  competing method.
	The last two rows represent the sagittal view.
	We intentionally ommited the coronal slicing due to it is the less informative given the properties
	  of the distortion.
	Red arrows point to regions where the accuracy of the proposed method more clearly overperformed
	  the competing method.
	}\label{fig:results_real}
\end{figure*}

### 2015-02-20 Publications/2015-NeuroImage/06-discussion.tex
discussion refactoring

% -*- root: 00-main.tex -*-
\section*{Discussion}
\label{sec:discussion}
We present a variational framework using to simultaneously segment and
  register 3D \gls*{dmri} of the human brain using the corresponding-anatomical image
  as reference.
The aproach is valid for image processing applications in which a precise segmentation is
  available (from an atlas or corresponding anatomical data) and match any or all
  the following circumstances:
  i) reference and target images do not present appropriate contrasts to
  	perform volume-based registration reliably;
  ii) the low resolution of the target volume produces strong partial volume effects
  	that exclude the use of voxel-wise segmentation algorithms and hinder volume-based
  	registration; and,
  iii) the estimation of the unknown warping is explicitly needed and the
    deformation field is smooth enough to be represented with a sufficient number
  	of B-Spline kernels (i.e. the distortion in \gls*{dmri} images,
  	object tracking in time series, etc.).

The algorithm is designed for the proposed application on correcting distortions of
  \gls*{dmri} data, a required preliminary step to extract the structural connectivity
  networks.
As a consequence, our approach relies on the availability of precise segmentations or
  surfaces of the objects that are to be segmented on the target volume(s).
Connectome extraction protocols generally include the acquisition of \gls*{t1} images
  to provide with prior information about anatomy with high accuracy.
Our method solves in a single step a joint problem usually addressed in two steps.
The first solves the linear registration of the \gls*{t1} and \gls*{dmri} data (i.e.
  \cite{greve_accurate_2009}), whereas the second corrects for the nonlinear distortions
  derived from the inhomogenety of magnetic susceptibility across tissues
  \citep{jezzard_correction_1995}.

In section \nameref{sec:methods_background}, we framed our method within the current
  state-of-art and defined the novelty of our approach.
The registration method performs the segmentation of the target image in several
  competing regions defined explicitly by their limiting surfaces.
We chose active contours without edges as we were interested in the homogeneous
  regions of diffusion properties within the brain.
Finally, a B-Spline deformation field supports and keeps track of the evolution of the
  active contours.
Iteratively, the active contours are updated following the instant shape gradients
  (see \nameref{app:shape_priors} in Appendix) with a gradient descent strategy.
To support our claim that images can be simultaneously segmented and registered to a
  reference coordinate system, we conducted evaluation experiments on digital
  simulated phantoms and real datasets.

\todo[inline]{Add sentence about comment RW\#1 of IPMI2013: incorporate knowledge of
the EPI distortions into regularization}

\paragraph*{Availability and reproducibility statement}
{\color{red} We also considered the reproducibility of results a design requirement.
Therefore, real data are fetched from a publicly available repository
  (the Human Connectome Project \citep{essen_human_2012}) and all the software
  involved in this paper is also publicly released.
\autoref{fig:evworkflows} describes the general structure of the workflow we implemented
  as evaluation instrument, using \emph{nipype} \citep{gorgolewski_nipype_2011} to ensure
  reproducibility.

We implemented the registration method upon the widely used Insight Registration and Segmentation
	Toolkit (\url{http://www.itk.org}) with the aim to release a useful and free software bundle.
The instrumentation framework is also distributed with the registration method,
  and has been implemented using \emph{nipype} \citep{gorgolewski_nipype_2011}.
The pipelines include the automatic generation of the ``cortex'' phantom.
The real datasets are publicly available under the \gls*{hcp} project.
Thus, all the experiments and results presented in this paper are intended to be
  fully reproducible.}

>>>>

% -*- root: 00-main.tex -*-
\section*{Discussion}
\label{sec:discussion}
In section \nameref{sec:methods_background}, we framed our method within the current
  state of art and defined the contributions in our approach.
We chose active contours without edges as we were interested in the homogeneous
  regions of diffusion properties within the brain.
Iteratively, the active contours are updated following the instant shape gradients
  (see \nameref{app:shape_priors} in Appendix) with a gradient descent strategy.
The aproach is valid for image processing applications in which a precise segmentation is
  available (from an atlas or within subject anatomical data), and match any or all
  the following circumstances:
  i) reference and target images do not present appropriate contrasts to
  	perform volume-based registration reliably;
  ii) the low resolution of the target volume produces strong partial volume effects
  	that exclude the use of voxel-wise segmentation algorithms and hinder volume-based
  	registration; and,
  iii) the estimation of the unknown warping is explicitly needed and the
    deformation field is smooth enough to be represented with a sufficient number
  	of B-Spline kernels (i.e. the distortion in \gls*{dmri} images,
  	object tracking in time series, etc.).
To support our claim that images can be simultaneously segmented and registered to a
  reference coordinate system, we conducted evaluation experiments on digital
  simulated phantoms and real datasets.

The algorithm is designed for the proposed application on correcting distortions of
  \gls*{dmri} data, a required preliminary step to extract the structural connectivity
  networks.
Connectome extraction protocols generally include the acquisition of \gls*{t1} images
  to provide with prior information about anatomy with high accuracy.
Our method solves in a single step a joint problem usually addressed in two steps.
The first solves the linear registration of the \gls*{t1} and \gls*{dmri} data
  \citep{greve_accurate_2009}, whereas the second corrects for the nonlinear distortions
  derived from the inhomogenety of magnetic susceptibility across tissues
  \citep{jezzard_correction_1995}.

\paragraph*{Prospects}
Future extensions or improvements of \emph{regseg} similar to the evolution of existing
  alternative methodologies.
Incorporating knowledge about the distortion initializing our method with the theory-based
  correction using fieldmaps would be the most profitable.
The main pitfall using fieldmap correction is the challenging problem of unwrapping the
  range of the phase map, as it is clipped in the range $(-\pi, \pi]$.
The most extended solution to this is \emph{prelude} \citep{jenkinson_fast_2003}, and some
  alternatives have been proposed \citep{daga_susceptibility_2014} to circumvent the
  inherent unreliability of the method.
The \emph{regseg} algorithm would naturally correct missregistered areas due to errors in
  phase unwrapping, and errors derived to the important smoothing processes that are
  typically to clean up field maps.
This integration has been recently applied in the \emph{b0}-to-\gls*{t2} approach
  by \cite{irfanoglu_susceptibility_2011}, studied along with the virtues of using
  nonregular B-Spline grids.
This is also a potential extension to our deformation model.

Finally, we are currently studying substitute features to build the energy functional on,
  enabling to perform \emph{regseg} directly on the raw \gls*{dmri} data.

### 2015-02-20 Publications/2015-NeuroImage/07-conclusion.tex
Update all the sections
discussion refactoring

% -*- root: 00-main.tex -*-
\section{Conclusion}
\label{sec:conclusion}
%
We propose and make publicly available a joint
  segmentation-registration method with a proof of concept
  on a digital phantom and an application on diffusion \gls*{mri}
  images of the human brain.
Starting with the Bayesian approach to image segmentation, we derive
  a cost function that models piecewise smooth regions with
  multivariate normal distributions.
This is, implicitly, an \acrlong*{acwe} model, in which the
  evolution of the surfaces drives the optimization of a B-Spline
  deformation model.

The phantom is used to demonstrate the success of the approach,
  on a quantitative evaluation that demonstrated its performance
  with sub-voxel precision (average vertex displacement close to
  $0.7mm$ in phantoms with isometric pixels of $1.0mm$ and $2.0mm$
  side).
We apply the method on the correction of susceptibility-derived
  distortions of diffusion \gls*{mri} data, and compare to
  a widely-used method based on the acquisition of extra images
  mapping the inhomogeneity of tissue susceptibility inside
  the scanner.

>>>>

% -*- root: 00-main.tex -*-
\section*{Conclusion}
\label{sec:conclusion}

\emph{Regseg} is a variational framework to simultaneously segment and
  register 3D \gls*{dmri} data of the human brain using within subject
  anatomical information as reference.
The registration method segments the target multivariate image in several competing regions
  defined explicitly by their limiting surfaces.
The surfaces are active, and evolve on a B-Spline displacements field, guided by shape
  gradients computed on the current partition of the target image.
\emph{Regseg} uses active contours without edges, and hence looks for
  homogeneous regions within the image.
We tested \emph{regseg} on digital phantoms simulating \gls*{t1} and \gls*{t2} \gls*{mri},
	warped with smooth random deformations.
The resulting average misplacement of the contours was significantly lower than the
  image resolution of the phantoms.

We propose \emph{regseg} to correct \gls*{dmri} datasets for the typically present distortions
  caused by inhomogeneitis of the main magnetic field.
Visual assessment of \emph{regseg} and cross-comparison with a widely-used competing
  technique demonstrated its accuracy, and overperformed the competing method.
Moreover, \emph{regseg} does not require additional images to the minimal acquisition protocol
  that generally includes only \gls*{t1} and \gls*{dmri}.

\section*{Availability and reproducibility statement}
\label{sec:availability}
We considered the reproducibility of our results a design requirement.
Therefore, we used real data fetched from a publicly available repository
  (the Human Connectome Project \citep{essen_human_2012}) and all the software
  involved in this paper is also publicly released.
The \emph{regseg} algorithm is implemented in C++, on top of ITK-4.6
  (Insight Registration and Segmentation Toolkit, \url{http://www.itk.org}).
The evaluation instruments (\autoref{fig:evworkflows}), are implemented in Python using
  \emph{nipype} \citep{gorgolewski_nipype_2011} to assess their reproducibility.
Phantom data are automatically generated by the evaluation framework.
All research elements (data, source code, figures, manuscript sources, etc.) in this paper
  are made publicly available under a unique digital object identifier in a
  package \citep{esteban_acweregistration_2015}.
